Does high personal relevance mitigate the effect of symbolic elements in government public communications?

### 2018-06-21 paper3_relevance and symbols_01.Rmd
Add files via upload

*Previous public administration studies have examined whether public communications can increase citizens trust in government organizations. Recent experimental studies have demonstrated that these communications can enhance trust not only by conveying substantive information, but also by incorporating strategically desigened symbolic element that evoke positive feelings and associations (e.g. logos, images and celebrities). Still, previous studies did not examine the conditions under which citizens are more/less susceptible to these symbols.

Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements can increase citizens' trust via two mechanisms: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by leading citizens to dedicate less attention to the substantive information in the communication, and thus causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust. I examine these expectations via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their undervaluation of public-service performance. One of the main remedies which has been proposed for this problem is greater investment in direct government communications with citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@deacon_2001; @eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

The public administration research on the effect of government public communications on citizens attitudes has focused mostly on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on organizations' performance and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

Recently, a few studies pointed to the additional effect of *symbolic* elements that are entangled in government communications, such as  agencies' names and brand logos, figures, images and celebrity endorsements [@alonbarkat_2017; @karens_2016; @teodoro_2018]. These studies demonstrated that government organizations may cause citizens to view them more favorably through familiar and well-designed symbolic elements that evoke positive associations and emotions. Still, we lack a theory and robust empirical evidence of the exact mechanisms through which these symbols increase citizens trust. Additionally, these previous studies did not examine the interaction between the symbolic elements and the substantive content of information in public communications.

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a normative view of democratic theory, the potential effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Mainly, we should be concerned about the possibility that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information of the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions).

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to think about its content. I discuss the theoretical and normative implications of these findings.

@@

# Symbolic elements in government communications and their shaping of citizens' attitudes: what do we know?

Symbols are fundamental for almost any form of human interaction, as acknowledged by early scholars of social science [@blumer_1986]. A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. My focus is on those symbols that are strategically used by government organizations in their public communications, and are designed to distinguish organizations and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can be also found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

By comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the name of US federal agencies not only represents information about their activities, but may also carry specific symbolic meaning and affect that can shape citizens' support for government policy (depending on citizens' political party identification). These two studies provide examples for the role of symbols in shaping citizens' attitudes. Still, they did not effectively disentangle the effects of the symbolic and informational aspects of the studied public communications.

A study by @karens_2016 did examine the distinct effect of symbols in communications on citizens' trust. The scholars surveyed university students from three European countries and examined their trust in two policy plans of the European Commission, and it's shaping by the appearance of the organization's logo in the communication of the policy plans. They found that including the Commission’s logo significantly increased citizens' trust in these policies. Their results show an impressive effect size equivalent to 0.7 standard deviations. [^footnote_karens]

[^footnote_karens]: The effect size they found was smaller with regard to the "air quality" policy compared with the "digital skills" policy which were presented (Cohen's d = 0.7 vs. 0.6). Also, the effect was smaller among subjects from the Netherlands, compared with those from Belgium and Poland (Cohen's d = 0.5 vs. 0.7).

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared between citizens who were personally exposed to prolonged power outages, and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the assessed government quality.

The aforementioned studies tend to suggest that public communications increase citizens' trust in government organizations and their policies, not only by conveying valuable substantive information, but also by incorporating symbolic elements that evoke positive associations and emotions. Moreover, the findings of @alonbarkat_2017 suggest that incorporating symbolic elements in public communications can unduly compensate for poor government performance, thus pointing to their potential detrimental implications.

Even though these studies substantiate a link between the appearance of symbolic elements in communications and citizens' favorable attitudes, they did not directly investigate the causal mechanisms that underlie this effect. Therefore, we still lack understanding of the circumstances under which symbols are likely to be more/less effective in shaping citizens' trust. Specifically, these studies neither theorized nor empirically tested the interaction between symbolic elements and citizens’ processing of substantive information in communications. We do not know whether well-designed symbols reinforce, or offset, citizens' rational responses to information. Specifically, as explained above, it is essential to assess the possibility that symbols can unduly compensate for unfavorable or logically unpersuasive information that should yield negative responses.

To address this theoretical gap in public administration research, I propose a nuanced theoretical framework for the effects of symbols in communications and their interaction with information, building on social psychology theory on persuasive communication.

# The Elaboration Likelihood Model and the various mechanisms of persuasion by elements in communications

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" communications [@petty_1986; @petty_2011]. In addition to its general theoretical application in psychological research, this model has been applied to studies on persuasion in various areas, including commercial advertising [@petty_1983; @swasy_1985; @dotson_2000; @drge_1989; @wheeler_2005; @moore_1987], public health campaigns [@wilson_2007; @angst_2009; @ruiter_2001; @briol_2006], media coverage [@petty_2002; @vanderwurff_2016] and media entertainment [@young_2008; @slater_2002]. Yet, thus far, the relevance of ELM has not been directly examined in relation to citizens' responses to government communications.

According to ELM, the responses of individuals to persuasive communications depends on the amount of thinking that they dedicate to the message content. Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little amount of cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion through the central route is expected to be more enduring, resistant and predictive of behavior, compared with peripheral route persuasion. Of course, much of the time, people are located in between the two ends of the continuum, and thus persuasion is determined by a mixture of these processes.

[^footnote_dual_process_theories]: The conceptual distinction between the "central and peripheral routes for persuasion" largely resembles the distinctions between heuristic and systematic processing [@chaiken_1980], and systems 1 and 2 [@kahneman_2011].

ELM enables to disentangle the distinct effects of different elements in communications. Some elements are processed differently by people at different locations along the elaboration continuum. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as arguments, and systematically scrutinize it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all. For example, the physical attractiveness of the source normally does not represent any substantive argument, yet it can serve as a simple cue. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

ELM further postulates that the placement of people along the elaboration continuum in any given situation is determined by their *motivation* and *ability* to think about the message. Accordingly, the literature proposes a variety of such motivational and ability factors. Among others, studies have demonstrated that the more people are distracted, the more difficult it is for them to scrutinize the message arguments, a mechanism that has been referred to as "elaboration disruption" [@petty_1976; @petty_1986]. Another prominent factor that influences people's inclination to elaborate on the message is their affective state, or mood. Positive mood (compared with neutral or negative moods) reduces people's cognitive ability and attenuates their motivation to think, due to their desire to maintain their positive mood and avoid message processing that would depress it [@worth_1987; @mackie_1989; @bless_1990; @schwarz_1991; @petty_2015; but see: @wegener_1995].[^footnote_elm_factors] Affecting the amount of thinking can cause attitude change in interaction with elements in the communication that can function as arguments or peripheral cues. Decreasing the amount of thinking weakens the effect the substantive arguments and enhances the effect of peripheral cues (and vice versa).

[^footnote_elm_factors]: For a thorough review of motivational and ability factors affecting likelihood to elaborate see @petty_1998.

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]; The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuades people directly through the peripheral route, yet at the same time they can also as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

[^footnote_multifunction_factors]: @petty_1986 have suggested that we are more likely to find complex relations of multiple mechanisms in those situations where people are located in the middle of the elaboration continuum. When people are neither extremely motivated or unmotivated to think, they may rely on different elements in the communications to decide if the message is worth or needs considering.

# ELM and the roles of symbolic elements and information in government public communications

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust through the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).

Symbolic elements may decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and absorbing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message - government organizations and policies - is likely to yield negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood. The latter expectation entails an interaction between symbols and the substantive information in the communications. Symbolic elements are likely to increase citizens' trust insofar as they reduce citizens' inclination to scrutinize information that would have decreased their trust.

@@

<br>
**Figure 1: Theoretical model**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/theoretical_model.png) 
<br>

Analytically, the expectation that symbolic elements decrease message elaboration also entails that symbols may decrease citizens' trust in communications about highly positive performance and persuasive policy plans. Nevertheless, one can make the argument that neutralizing the effect of "positive" information with symbols may not actually yield significant differences, since citizens are more likely to "punish" bureaucracies for publishing negative information than to "praise" them for publishing positive information [@james_2011; @nielsen_2017; @james_2017b]. Congruently, the effect of positive information (versus a medium level) is likely to be relatively small, and hence neutralizing this effect may not make much difference. Given the focus of this study on the interaction between symbols and ill-conceived policies, the research design does not directly addresses this latter interaction.

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans. Such communications can be found in publicized reports, pamphlets, websites, new media pages and ads. Focusing on communication of future policy plans, as opposed to communications on performance of existing policies, enables me to utilize a real-world empirical case without compromising the reliability  of the manipulation of the content. This is because most citizens do not have prior knowledge of organizations' future policy plans, and thus they will be more willing to trust a fictional communication of an organization with regard to its future policy.

In addition to the above-mentioned hypotheses, the experiment is also designed to test an additional set of hypotheses regarding the interaction between symbols and the perceived personal relevance of communications. The selection of policy plans and the construction of the population sample were designed to account for variation in that variable. In this paper, I focus on those methodological and empirical parts of the study that are consequential for the examination of the interaction between symbols and information.[^footnote_relevance_paper]

[^footnote_relevance_paper]: The methodological aspects which are relevant for examining the additional hypotheses (i.e. the interaction between symbols with personal relevance) are described in a separate paper. An English translation of the full survey is available from the supplementary appendix.

@@

Empirically, I focused on the case of the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, both at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM makes a good case for this study since has invested greatly over the past years in public communications. In these various communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (image a. in figure 1). This logo consists of pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element in which I focused is the green color, which is strongly associated with taking care of the environment. The EPM's logo and the green color can be seen in almost all of the visual communications of the EPM, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, that previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (image b.) was launched in 2010, and included a series of advertisements in which Friedman was presented with a green grass head, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (image c.) was launched in 2017, and was focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

In the course of the survey experiment, I confirmed that the EPA logo and the images of the celebrities from the campaigns are familiar to the participants. `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize EPA logo to some extent and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`% , respectively). In robust analyses I further exclude those who do not recognize these symbols.[^footnote_prior_studies]

[^footnote_campaigns]: Multimedia files of these campaigns are available in EPM's website: http://www.sviva.gov.il/InfoServices/ReservoirInfo/ResearchAndPublications/Pages/default.aspx.

[^footnote_prior_studies]: In two prior online survey tests, which I conducted before the experiment, I collected further information about the affect and symbolic associations of these symbols. These results, which are reported in the supplementary appendix, substantiate my empirical assumption that these elements evoke positive emotions Israeli citizens, and that citizens automatically associate them with EPM and its policies. These results also indicate that the images of the celebrity comedians from EPM campaigns are perceived as humoristic, and accordingly, they are expected to position people in a positive mood.

<br>
**Figure 2: symbolic elements of EPM**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/EPM_symbols.png) 
<font size="0.5">

Notes:

a. EPM logo.
b. Tal Friedman, comedian,  EPM campaign “starting to think green” (2010): "Isn't money growing on trees? Moving to environmental consumption and saving at least 6,000 NIS per year!", "starting to think green", "The Environmental Protection Ministry".
c. Ido Rosenblum, comedian, EPM campaign  “taking every bag seriously” (2017): "The reusable bag is again left at home? A tip from Ido: Hang it on the door handle and you will not leave without it", "Taking every bag seriously", "The Environmental Protection Ministry".

@@

## *Procedure and sample population*

The survey was presented to the participants as a study about citizens' attitudes regarding environmental policy issues. The participants were first asked pre-manipulation questions about their political ideology and their perceptions about the pubic sector.[^footnote_relevance_manipulation] Thereafter, they were told that they will be presented with two government policy plans regarding specific environmental issues, obtained from EPM's official publication of its annual work plan for the year 2018. The policy plans regarded to reducing air-pollution in the Bay of Haifa (an area in Israel) and reducing domestic waste and increasing recycling. The two policy plans were presented to each subject in a random order. Each policy plan included the title of the policy goal (i.e. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

[^footnote_relevance_manipulation]: This section also included a manipulation of participants' perceived personal relevance. Participants in the treatment group were asked a series of questions about environmental government policy, whereas participants in the control group were asked instead equivalent questions about their field of occupation.

Per each policy, the subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). The policy activities which were mentioned in the strong policy plans condition were taken from the real EPM work plan and from other reports and official sources of EPM. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures which are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were increasing the supervision of polluting factories, and raising citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is displayed in **Appendix A**. For each subject, one of the two policies was strong and the other was weak. For example, a subject who first received a strong air-pollution policy, later received the weak recycling policy, and vice versa.

In addition to information manipulation, the participants are randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols): treatment (real symbols), control no symbols and control fake symbols. In the symbol treatment condition, the two policies (again, one weak and one strong) are displayed with the above-mentioned three symbolic elements: the titles and the footer are colored with green, EPM logo is displayed at the top of the page, and the celebrities' images from the campaigns are attached and labeled "EPM advertisement". Each communication included the image of one celebrity: Rosenblum's image from the campaign for reducing the use of disposable bags is attached to the recycling policy; Friedman's image from the "starting to think green" campaign was attached to the air-pollution policy. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities images. Finally, in the control fake-symbols condition, the green color is replaced with blue, the EPM logo is replaced by a fake logo and the images of celebrities from the campaigns are replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition is to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols to that they will resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which can be attached to the government ministry and environmental issues.[^footnote_symbols_manipulation_checks] In that sense, it can be regarded as a placebo. The symbols conditions are displayed in **Appendix B**. Further details about the images in the real and fake symbols conditions are found in the supplementary appendix. That each subject received two policies which were presented under the same symbol condition, but under different information conditions (i.e. one strong and one weak), enables me later to analyze the interaction between these two manipulations both via between- and within-subjects.

[^footnote_symbols_manipulation_checks]: In two prior surveys, I empirically confirmed that the fake symbols are unfamiliar, do not have strong positive or negative affect, and are not strongly associated with environment policy and EPM. The results of these examinations are reported in the supplementary appendix.

Each policy plan was followed by six questions questions about respondents' trust in the policy. Afterwards, the subjects conducted a memory test on the last policy that was presented to them, and were asked about the extent to which they elaborated on it and about the personal relevance of the policy plans. Thereafter, I included an instructional manipulation check (IMC) item, which aims to assess participants’ diligence. After a set of demographic questions, I conducted a manipulation check, in which I asked the participants whether they recognize the Ministry's logo, and the two images of the celebrities. Finally, the participants were debriefed about the manipulations, and the possible deception. An English translation of the full survey is available from the supplementary appendix.

Altogether, for each policy plan, the survey experiment has a factorial design of 3 (symbols: real symbols, fake-symbols, no-symbols) x 2 (information: strong vs. weak policy). The assignment to the different experimental groups is graphically presented in **Figure 3**. The experimental groups are fairly balanced with respect to demographics, and pre-manipulation variables. Throughout the rest of the paper, I will analyze the results at the policy level, in order to simplify the analyses. Accordingly, I transformed the dataset into a multilevel structure of policies nested within subjects.[^footnote_combined_analyses]

[^footnote_combined_analyses]: In the supplementary appendix I further replicate the analyses below, while adding the interactions between the manipulations and the policy plans. In these models, the interactions between the policy plans and the symbols are far from significance, with no major changes to the results.

<br>
**Figure 3: Experimental design for policy plan communications**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/experiment_flow.png)

The survey experiment was conducted online through Qualtrics Survey Software. To subjects, adult Israeli citizens, were recruited via an Israeli internet research panel company - *iPanel*. The online survey link was sent by iPanel, between 28.1-1.2.2018 to 5,077 people. A total of `r N_0` respondents completed the survey (a response rate of `r round((N_0/5077)*100,1)`%).[^footnote_responserate] Of these, I later filtered out `r N_0-N_1` observations (`r round(((N_0-N_1)/N_0)*100,1)`%) due to multiple entries from the same IP address (n=`r filter_IP`), failure at the instructional manipulation test (n=`r filter_IMC`), submitting the survey in less than 3 or more than 30 minutes (n=`r filter_time`), and respondents under the age of 18 (n=`r filter_age`). These filters are not significantly associated with the experimental conditions. Ultimately, following this rigorous screening, I ended up with a sizable sample of `r N_1` subjects (Mean age = `r SVIVA2_01%>%summarise(mean(AGE,na.rm=T))%>%max()%>%round(1)`, `r SVIVA2_01%>%summarise(mean(GENDER,na.rm=T)*100)%>%max()%>%round(1)`% women).[^footnote_sample_areas] Because each subjects evaluated two different policy plans, my dataset has a multilevel structure of `r SVIVA2_01_comb%>%nrow()` observations nested within these `r N_1` subjects.

[^footnote_responserate]: According to iPanel's reporting, 330 women entered the survey, but were eventually denied of access, to avoid the over-representation of women in the sample. Hence, the response rate, when reducing those who were denied of access, is slightly higher  and reaches `r round((N_0)/(5077-330)*100,1)`%.

[^footnote_sample_areas]: To capture variation in the personal relevance of the air-pollution policy, I sampled citizens from the major cities in the Haifa-Bay area, and matched them with citizens from cities with a similar profile and population size at the center of the country. The final sample consists of `r SVIVA2_01_haifa%>%nrow()` subjects from the Haifa-bay cities, and `r SVIVA2_01_center%>%nrow()` from the center cities. The two areas received the exact same link to the online survey, and they were equally, randomly assigned to the experimental conditions.

##*Operationalization of variables*

```{r, echo=FALSE, warning=FALSE, message=F}
alpha_air_trust = psych::alpha(select(SVIVA2_01, TRUST_air_q1,
                               TRUST_air_q2,
                               TRUST_air_q3,
                               TRUST_air_q4,
                               TRUST_air_q5,
                               TRUST_air_q6))[["total"]][["raw_alpha"]]%>%round(2)

alpha_waste_trust = psych::alpha(select(SVIVA2_01, TRUST_waste_q1,
                               TRUST_waste_q2,
                               TRUST_waste_q3,
                               TRUST_waste_q4,
                               TRUST_waste_q5,
                               TRUST_waste_q6))[["total"]][["raw_alpha"]]%>%round(2)

alpha_trust_comb = ((alpha_air_trust+alpha_waste_trust)/2)%>%round(2)

cor_elaboration_air = cor.test(SVIVA2_01$MEMORY_air_score,SVIVA2_01$ELABORATION_air_time_log,method="spearman")$estimate%>%mean()%>%round(2)

cor_elaboration_waste = cor.test(SVIVA2_01$MEMORY_waste_score,SVIVA2_01$ELABORATION_waste_time_log,method="spearman")$estimate%>%mean()%>%round(2)

cor_RT_delta = cor.test(SVIVA2_01$ELABORATION_total_time_log,
                          SVIVA2_01$WITHIN_DELTA)$estimate%>%mean()%>%round(2)

cor_memory_delta = cor.test(SVIVA2_01$MEMORY_total,
                          SVIVA2_01$WITHIN_DELTA)$estimate%>%mean()%>%round(2)

>>>>

*Previous public administration studies have examined whether public communications can affect citizens trust in government organizations and their policies. Recent experimental studies have demonstrated that these communications can enhance trust not only by conveying substantive information, but also by incorporating strategically desigened symbolic element that evoke positive feelings and associations (e.g. logos, images and celebrities). Still, under which conditions are citizens more/less likely to be emotionally persuaded by these symbols? Building on the social psychology Elaboration Likelihood Model (ELM), I theorize that citizens are less susceptible to symbols when they perceive the policy issue in the communication as having personal relevance for themselves, which enhances their motivation to invest in elaborate processing. I put this theory to test through a survey experiment, where I examine circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens' trust in the Environment Protection Ministry's policies for reducing the air pollution in a specific area (the Haifa-Bay) and reducing domestic waste. I operationalize personal relevance both experimentally, and by comparing between citizens from the polluted Haifa-Bay area and others. The experimental results do not support the hypothesis that higher personal relevance attenuates the effect of symbols. They tentatively suggest that symbols are more likely to persuade citizens who have predominant negative attitudes.*

@@

Public administration scholars have explored whether government organizations can enhance citizens' trust in them by employing public communications. Whereas most of the studies on the effect of government public communications on citizens attitudes has focused on the *informative* aspects of these communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @grimmelikhuijsen_2015; @ho_2017], recent experimental studies further pointed to the emotive effect of *symbolic* elements that are entangled in government communications, such as  agencies' names [@teodoro_2018], brand logos [@karens_2016], and figures in public campaigns [@alonbarkat_2017]. These studies demonstrated that these familiar and well-designed symbolic elements can enhance citizens' trust irratioanlly through a mechanism of "evaluative conditioning" - unconscious transfer of affect from the symbols to the public organizations and policies associated with them [@dehouwer_2012].

While the abovementioned studies have substantiated the distinct effect of symbols, they did not theorize, nor empirically examine the conditions under which are these symbols are likely to be more/less effective in persuading citizens. The research gap has great significance. From the organizations' practical point of view, ...

Employing a broader, democratic normative stance, we should be also concerned from the effect of symbols. Symbolic elements can elicit undue trust in poorly performing organizations and in ill-conceived policies. Accordingly, it may enable public organizations to escape justified public criticism [@alonbarkat_2017]. Hence, by understanding when citizens are more likely to be affected by symbols, we  may enable us to identify the

Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *under what conditions do symbolic elements  increase citizens' trust in governemnt policy?* And specifically, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. I expect that this effect is attenuated the more citizens perceive the communications as personally relevant, because they are more motivated to think about their messages. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard negative of logically unpersuasive content that would have otherwise decreased their trust.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow() %>% round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry, and specifically the extent to which their trust in these policies are affected by the substantive information of the communications (logically persuasive vs. unpersuasive policy plans) and by symbolic elements in them (real symbols vs. fake symbols and no symbols conditions). In addition, I examine the moderating role of the perceived personal relevance of these policies.

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, who were distracted by the symbols and thus were less inclined to think about its content. I did not find support for the hypothesis that the effect of the symbolic elements is attenuated by the perceived personal relevance of the policy issue. I discuss the theoretical and normative implications of these findings.

@@

# The Elaboration Likelihood Model (ELM) and the effect of symbols and information in communicaions.

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" 
<br>

# The moderating role of perceived personal relevance on citizens' responses to symbols and information in communications

@@

@@

<br>
**Figure 1: Theoretical model**

@@

@@

@@

##*Data collection*

## *Procedure*

<br>
**Figure 3: Experimental design for policy plan communications**
<br>

##*Variables*
